#### Day 1 Challenges

If you're starting a new session to answer these questions, you'll need to read in the wetland and tree data frames again.

<p class = 'code'>Read in example ACAD wetland data from url</p>
```{r c1rdat, eval = F}
ACAD_wetland <- read.csv(
  "https://raw.githubusercontent.com/KateMMiller/IMD_R_Training_2026/refs/heads/main/data/ACAD_wetland_data_clean.csv"
  )
```

<p class = 'code'>Read in example NETN tree data from url</p>
```{r c1readtree, eval = F}
trees <- read.csv("https://raw.githubusercontent.com/KateMMiller/IMD_R_Training_2026/refs/heads/main/data/NETN_tree_data.csv")
```


<div class="alert alert-info">
<strong>Question 1. How would you look at the the first 4 even rows (2, 4, 6, 8), and first 2 columns?</strong>

<details><summary class='code2'>Answer</summary>
```{r d1chal1, fold.output = F}
ACAD_wetland[c(2, 4, 6, 8), c(1, 2)]
```
</details>
</div>
<br>

<div class="alert alert-info">
<strong>Question 2. How many unique species are there in the ACAD_wetland data frame?</strong>

<details><summary class='code2'>Answer</summary>
```{r d1rows2a, results = "hide"}
# Option 1
length(unique(ACAD_wetland[, "Latin_Name"]))
```

```{r d1rows2b, fold.output = FALSE}
# Option 2
length(unique(ACAD_wetland$Latin_Name)) # equivalent
```
</details>
</div>
<br>

<div class="alert alert-info">
<strong>Question 3. Which sites have species that are considered protected on them (Protected = TRUE)?</strong>

<details><summary class='code2'>Answer</summary>
```{r, d1rows2a2, results = 'hide'}
# Option 1 - used unique to just return unique site name
unique(ACAD_wetland$Site_Name[ACAD_wetland$Protected == TRUE])

```

```{r, d1rows2b2, fold.output=FALSE}
# Option 2
unique(ACAD_wetland[ACAD_wetland$Protected == TRUE, "Site_Name"])
```
</details>
</div>
<br>

<div class="alert alert-info">
<strong>Question 4. How many trees are on Plot MIMA-012?</strong>
<details><summary class='code2'>Answer</summary>



```{r d1chal4}
mima12 <- subset(trees, Plot_Name == "MIMA-012")
nrow(mima12) # 12
```
</details>
</div>
<br>

<div class="alert alert-info">
<strong>Question 5. How many trees with a TreeStatusCode of "AS" (alive standing) are on Plot MIMA-012?</strong>
<details><summary class='code2'>Answer</summary>
```{r d1chal5, fold.output = F}
# OPTION 1
mima12_as <- subset(trees, Plot_Name == "MIMA-012" & TreeStatusCode == "AS")
nrow(mima12_as) # 6

# OPTION 2
mima12 <- trees[trees$Plot_Name == "MIMA-012",]
table(mima12$TreeStatusCode) # 6

```
</details>
</div>
<br>

<div class="alert alert-info">
<strong>Question 6. Find the DBH record that's > 400cm DBH.</strong>

<details><summary class='code2'>Answer</summary>
There are multiple ways to do this. Two examples are below.

Option 1. View the data and sort by DBH.
```{r d1vtrees, eval = F}
View(trees)
```

Option 2. Find the max DBH value and subset the data frame
```{r d1maxdbh, fold.output = F}
max_dbh <- max(trees$DBHcm, na.rm = TRUE)
trees[trees$DBHcm == max_dbh,]
```
</details>
</div>
<br>

<div class="alert alert-info">
<strong>Question 7. What is the value of the largest DBH, and which record does it belong to?</strong>

<details><summary class='code2'>Answer</summary>
There are multiple ways to do this. Two examples are below.

Option 1. View the data and sort by DBH.
```{r d1vtrees2, eval = F}
View(trees)
```

Option 2. Find the max DBH value and subset the data frame
```{r d1maxdbh2}
max_dbh <- max(trees$DBHcm, na.rm = TRUE)
max_dbh #443
trees[trees$DBHcm == max_dbh,]
# Plot MIMA-016, TagCode = 1.
```
</details>
</div>
<br>

<div class="alert alert-info">
<strong>Question 8. How would you check that the line of code above worked?</strong>

<details><summary class='code2'>Answer</summary>
There are multiple ways to do this. Two examples are below.

Option 1. Show the range of the original and fixed data frames
```{r d1check1, fold.output = F}
range(trees$DBHcm)
range(trees_fix$DBHcm)
```

Option 2. Plot a histogram of the original and fixed data frames
```{r d1check2, fold.output = F}
hist(trees$DBHcm)
hist(trees_fix$DBHcm)
```

Option 3. Calculate max of DBHcm column
```{r d1check3, fold.output = F}
max(trees$DBHcm)
max(trees_fix$DBHcm)
```
</details>
</div>
<br>