
#### Basic Plotting
Visualizing the data is also important to get a sense for the data and look for potential errors and outliers. 

Histograms are a great start. The code below generates a basic histogram plot of a specific column in the dataframe using the `hist()` function.

```{r readtree2, results = 'hide', echo = F}
trees <- read.csv("https://raw.githubusercontent.com/KateMMiller/IMD_R_Training_2026/refs/heads/main/data/NETN_tree_data.csv")
```


<p class='code'>Plot histogram of DBH measurements</p>
```{r hist, fold.plot = F, fig.height = 4, fig.width = 6}
hist(x = trees$DBHcm)
```
<br>
Looking at the histogram, it looks like all of the measurements are below 100cm except for one that's way out in 400 range. You can also make a scatterplot of the data. If you only specify one column, the x axis will be the row number for each record, and the y axis will be the specified column.

<p class='code'>Make point plot of DBH measurements</p>
```{r pdbh1, fold.plot = F, fig.height = 4, fig.width = 6}
plot(trees$DBHcm)
```
<br>
Again, you can see there's one value that's greater than all of the others. 

We can also plot two variables in a scatterplot. 

<p class='code'>Make scatterplot of crown class vs. DBH measurements (Option 1)</p>
```{r pdbh2, fold.plot = F, fig.height = 4, fig.width = 6}
plot(trees$DBHcm ~ trees$CrownClassCode)
```

<p class='code'>Make scatterplot of crown class vs. DBH measurements (Option 2- better axis labels)</p>
```{r pdbh5, fold.plot = F, fig.height = 4, fig.width = 6}
plot(DBHcm ~ CrownClassCode, data = trees) # equivalent but cleaner axis titles

```
<br>
Again, you can see there's one value that's greater than all of the others, and it's crown class code 3 (codominant). 

<div class="alert alert-info">
<strong>Question: Plot a histogram of percent cover (`Ave_Cov`) in the `ACAD_wetland` </strong>

<details><summary class='code2'>Answer</summary>
```{r wethist, fig.height = 4, fig.width = 6}
hist(ACAD_wetland$Ave_Cov)
```
</details>
</div>
<br>
