---
output: 
  html_document:
    css: custom_styles.css
---
```{r}
knitr::opts_chunk$set(warning=FALSE, message=FALSE, fig.align = 'center', fig.height = 3, fig.width = 5)
```

#### Intro to `ggplot2`
<details open><summary class='drop'>Intro to `ggplot2`</summary>
The `ggplot2` package is the most popular `R` package for plotting. It takes a little effort to learn how the pieces of a ggplot object fit together. However, once you get the hang of it, you can create and customize a large variety of attractive plots with just a few lines of `R` code. The package is called `ggplot2` because originally there was `ggplot`. The developer, Hadley Wickham, didn't want to break the original package to improve the package, so created ggplot2.  

[The ggplot2 online book]("https://ggplot2-book.org/") and [cheatsheets](https://posit.co/wp-content/uploads/2022/10/data-visualization-1.pdf") can be very helpful while you are learning to use the `ggplot2` package.

The `ggplot2` package was developed using the grammar of graphics as the underlying philosophy, which basically breaks a plot up into individual building blocks related to aesthetics (e.g., color, size, shape), geometries (e.g. points, lines, boxes), and themes (e.g. axis label font size, legend placement, etc.).

Important concepts with `ggplot2`:
<ul>
<li><b>`ggplot(data, aes())`:</b> Every ggplot object starts with this line, which tells R the data you're plotting, and which variables you're plotting where in `aes()` argument.</li>
<li><b>data: </b> every plot requires data. The first argument for every `ggplot()`call is the data. This also means you can pipe data into a ggplot object. </li>
<li><b>aes: </b> Short for aesthetics. This is where you tell ggplot what your x and y variables are. If you want aesthetics, like color, fill, or size to vary by the data (e.g., color code a figure by park, use a dashed vs. solid line to distinguish between significant/non-significant trend, etc.), those variables are specified within the `aes()` argument either at the `ggplot()` level, or within the specific geom. </li>
<li><b>geom: </b> Short for geometries, geoms represent what you see in the plot, such as the points in a scatter plot, the lines in a trend plot, or the boxes in a boxplot.</li>
<li><b>scale: </b>Scales go hand in hand with `aes()`. If you specify `aes(color = park)`, then scale is where you can specify a custom color for each park instead of ggplot's default color scheme. The scale is where you can set the labels of groups in a legend (if different than how the data are labeled). You can also customize axis ranges, breaks and labels with different scales. </li>
<li><b>theme: </b>This is where you can change the format of the figure, such as removing the gridlines in the default ggplot format, making axis labels larger, changing the position of the legend, etc.</li>
<li><b>facet: </b>Faceting the data allows you to graph multiple plots based on a grouping variable (e.g., site, species, year, etc.) in the same ggplot object.</li>
</ul>

</details>
<br>

<details open><summary class='drop'>Park Visitation vs. Gas prices</summary>
We will build our first ggplot object step-by-step to demonstrate how each component contributes to the final plot. Our first ggplot object will create a line graph of trends in visits to Acadia National Park from 1994 to 2024.

<p class = 'code'>Step 1. Import the visitation and gas prices data</p>
```{r import_gvdata, echo = TRUE, eval = F, message = FALSE, warning = FALSE}
visit_vs_gas <- read.csv("https://raw.githubusercontent.com/KateMMiller/IMD_R_Training_2026/refs/heads/main/data/ACAD_Visits_vs_Gas_prices.csv")
```
```{r import_gvdata2, echo = F, results = 'hide', eval = T, message = FALSE, warning = FALSE}
visit_vs_gas <- read.csv("./data/ACAD_Visits_vs_Gas_prices.csv")
```

<p class = 'code'>Step 2. Look over the data for any potential problems.</p>

```{r view_data, eval = F, message = FALSE, warning = FALSE}
# Examine the data to understand data structure, data types, and potential problems
head(visit_vs_gas) 
summary(visit_vs_gas)
table(visit_vs_gas$Year) 
table(complete.cases(visit_vs_gas))
```
```{r view_data2, fold.output = F}
str(visit_vs_gas)
```

<p class = 'code'>Step 3. Fix the annual visitation variable</p>
Note that Annual visits is treated like a character because there are "," thousands separators. We have to fix that before we can plot the data. There are multiple ways to do this. I'm going to use the `gsub()` function that behaves like `gsub([find "pattern"], [replace with "pattern"], [column to search])`. The empty "" removes the ",". The `as.numeric()` converts a character to numeric. As long as you don't get the "NAs introduced through coersion" warning in the console, all rows in the column of interest were successfully. converted to a number.

```{r fixav}
# Base R
visit_vs_gas$Annual_Visits <- as.numeric(gsub(",", "", visit_vs_gas$Annual_Visits))

# Tidyverse
visit_vs_gas <- visit_vs_gas |> mutate(Annual_Visits = as.numeric(gsub(",", "", Annual_Visits)))
str(visit_vs_gas$Annual_Visits) #check that it worked
```

<p class = 'code'>Step 4. Create the ggplot template of annual visitation per 1000 visitors over time</p>
```{r p1, fold.plot = F, fig.height = 3, fig.width = 5, fig.align = 'center'}
p <- ggplot(data = visit_vs_gas, aes(x = Year, y = Annual_Visits/1000))
p
```

<p class = 'code'>Step 5. Add point geometry</p>
```{r p2, fold.plot = F, fig.height = 3, fig.width = 5, fig.align = 'center'}
p1 <- p + geom_point()
p1
```

<p class = 'code'>Step 6. Add line graph geometry</p>
```{r p3, fold.plot = F, fig.height = 3, fig.width = 5, fig.align = 'center'}
p2 <- p1 + geom_line()
p2
```

<p class = 'code'>Step 7. Fine tune the Y and X axis breaks</p>
```{r p4, fold.plot = F, fig.height = 3, fig.width = 5, fig.align = 'center'}
p3 <- p2 + scale_y_continuous(name = "Annual visitors in 1000's",
                              limits = c(2000, 4500),
                              breaks = seq(2000, 4500, by = 500)) + # label axis at 2000, 2500, 3000,... up to 4500
           scale_x_continuous(limits = c(1994, 2024),
                              breaks = c(seq(1994, 2024, by = 5))) # label axis at 1994, 1999, 2004,... up to 2024
p3
```

<p class = 'code'>Step 8. Change the labels</p>
```{r p5, fold.plot = F, fig.height = 3, fig.width = 5, fig.align = 'center'}
p4 <- p3 + labs(x = "Year", 
                title = "Annual visitation/1000 people in Acadia National Park 1994 - 2024")
p4
```
<br>
Note how the axis labels can be specified either in the scale (step 4) or in the `labs()` function.

<p class = 'code'>Step 9. Modify the theme</p>
There are built in themes that change the default formatting of the plot. Here I show `theme_bw()`, but there are many options. Play around with the other themes to get a feel for the options: `theme_linedraw()`, `theme_light()`, `theme_dark()`, `theme_minimal()`, `theme_classic()`, `theme_void()`. The two I use most often are `theme_bw()` and `theme_classic()`.  There's a package called `ggthemes` that you can install with even more themes. You can also create your own themes.

```{r p6, fold.plot = F, fig.height = 3, fig.width = 5, fig.align = 'center'}
p5 <- p4 + theme_bw() 
p5
```


#++++ ENDED HERE ++++
- Turn off gridlines manually
- add another plot where group, color, facets are used.

</details>

