#### Day 2 Questions

If you're starting a new session to answer these questions, you'll need to load dplyr and read in the tree data frame again.

<p class = 'code'>Load dplyr</p>
```{r ldpd2c}
library(dplyr)
```

<p class = 'code'>Read in example NETN tree data from url</p>
```{r c1readtree2c, eval = F}
trees <- read.csv("https://raw.githubusercontent.com/KateMMiller/IMD_R_Training_2026/refs/heads/main/data/NETN_tree_data.csv")
```

<p class = 'code'>Create the tree_final data frame</p>
```{r tfc22c}
trees_final <- trees |> 
  mutate(DecayClassCode_num = as.numeric(DecayClassCode),
         Plot_Name = paste(ParkUnit, PlotCode, sep = "-"),
         Date = as.Date(SampleDate, format = "%m/%d/%Y")) |> 
  rename("Species" = "ScientificName") |> 
  filter(IsQAQC == FALSE) |> 
  select(-DecayClassCode)
```

<h5>Data Wrangling with dplyr </h5>
<div class="alert alert-info">
<strong>CHALLENGE: How many trees are on Plot MIMA-12 (using trees_final)?</strong>
<details><summary class='drop2'>Answer</summary>
```{r chal4_d22c}
trees_final |> filter(Plot_Name == "MIMA-12") |> nrow()
```
</div>
</details>

<div class="alert alert-info">
<strong>CHALLENGE: How many trees with a TreeStatusCode of "AS" (alive standing) are on Plot MIMA-12 (using trees_final)?</strong>
<details><summary class='drop2'>Answer</summary>

```{r chal5d22c, fold.output = F}
trees_final |> filter(Plot_Name == "MIMA-12" & TreeStatusCode == "AS") |> nrow()
```

</details>
</div>

<div class="alert alert-info">
<strong>CHALLENGE: What is the exact value of the largest DBH, and which record does it belong to?</strong>

<details><summary class='drop2'>Answer</summary>
```{r maxdbhd22c, fold.output = T}
# Base R and dplyr combo
max_dbh <- max(trees_final$DBHcm, na.rm = TRUE)
trees_final |> 
  filter(DBHcm == max_dbh) |> 
  select(Plot_Name, SampleYear, TagCode, Species, DBHcm)

# dplyr with slice
trees_final |> 
  arrange(desc(DBHcm)) |> # arrange DBHcm high to low via desc()
  slice(1) |> # slice the top record
  select(Plot_Name, SampleYear, TagCode, Species, DBHcm)

```
</details>
</div>

<div class="alert alert-info">
<strong>CHALLENGE: Fix the DBH typo by replacing 443.0 with 44.3.</strong>

<details><summary class = 'drop2'>Answer</summary>
```{r trfxd22c, fold.output = F}
# Base R
# create copy of trees data
trees_fix <- trees3
# find the problematic DBH value, and change it to 44.3
trees_fix$DBHcm[trees_fix$Plot_Name == "MIMA-016" & trees_fix$TagCode == 1 & trees_fix$DBHcm == "443"] <- 44.3

# dplyr via replace
trees_fix <- trees |> mutate(DBHcm = replace(DBHcm, DBHcm == 443.0, 44.3))
```

Check that it worked by showing the range of the original and fixed data frames.
```{r check1d22c, fold.output = F}
range(trees$DBHcm)
range(trees_fix$DBHcm)
```
</details>
</div>
