#### Day 3 Questions
<details open><summary class='drop'>Load Data and Packages</summary>
```{r}
library(dplyr)
library(stringr) # for word 
bat_cap <- read.csv("https://raw.githubusercontent.com/KateMMiller/IMD_R_Training_2026/refs/heads/main/data/example_bat_capture_data.csv")
spp_tbl <- read.csv("https://raw.githubusercontent.com/KateMMiller/IMD_R_Training_2026/refs/heads/main/data/NETN_tree_species_table.csv")
trees <- read.csv("https://raw.githubusercontent.com/KateMMiller/IMD_R_Training_2026/refs/heads/main/data/NETN_tree_data.csv")
```
</details><br>

<details open><summary class='drop'>Pivoting tables</summary>
<div class="alert alert-info">
<strong>Pivot the `bat_sum` data frame on year instead of species, so that you have a column for every year of captures. Remember to avoid column names starting with a number. </strong>

<details><summary class='code2'>Answer</summary>
```{r pivwide2c1, fold.output = F}
bat_wide_yr <- pivot_wider(bat_sum, names_from = Year, 
                           values_from = num_indiv, 
                           values_fill = 0, 
                           names_prefix = "yr")
head(bat_wide_yr)
```
</details>
</div>

<div class="alert alert-info">
<strong>Pivot the resulting data frame from the previous question to long on the years columns, and remove the "yr" from the year names using `names_prefix = 'yr'`. </strong>

<details><summary class='code2'>Answer</summary>
```{r pivwide2c2, fold.output = F}
bat_long_yr <- pivot_longer(bat_wide_yr, 
                            cols = -c(Site, sppcode),
                            names_to = "Year", 
                            values_to = "num_indiv", 
                            names_prefix = "yr") # drops this string from values
```
</details>
</div>
</details><br> 

<details open><summary class='drop'>Joining Tables</summary>
<div class="alert alert-info">
<strong>Join the NETN_tree_data.csv and NETN_tree_species_table.csv to connect the common name to the tree data.</strong>

<details><summary class='drop2'>Answer</summary>
```{r challtr2bc1}
# find the columns in common
intersect(names(spp_tbl), names(trees)) # TSN and ScientificName
# left join species to trees, because don't want to include species not found in tree data
trees_spp <- left_join(trees, 
                       spp_tbl |> select(TSN, ScientificName, CommonName), 
                       by = c("TSN", "ScientificName"))

head(trees_spp)
```
</details>
</div>

<div class="alert alert-info">
<strong>Find a species in the NETN tree data that doesn't have a match in the species table.</strong>
<details><summary class='drop2'>Answer</summary>
```{r challtr2d}
# find the columns in common
intersect(names(spp_tbl), names(trees)) # TSN and ScientificName
# anti join of trees against species table, selecting only columns of interest
anti_join(trees, spp_tbl, by = c("TSN", "ScientificName")) |> 
  select(ParkUnit, PlotCode, SampleYear, ScientificName)
```
</details>
</div>

</details>
</br>

<details open><summary class='drop'>Dates and Times</summary>
<div class="alert alert-info">
<strong>Question: How would you return date1 as YYYYMMDD (20260312)?</strong>
<details><summary class='drop2'>Answer</summary>
```{r chald31dc}
format(date1, format = "%Y%m%d")
```
</details>
</div>

<div class="alert alert-info">
<strong>How would you create a list of dates in 2026 that are evenly spaced by 3 months?</strong>
<details><summary class='drop2'>Answer</summary>
```{r chald32bc}
date_list <- as.Date(c("01/01/2026", "12/31/2026"), format = "%m/%d/%Y") 
seq.Date(date_list[1], date_list[2], by = "3 months")
```
</details>
</div>

<div class="alert alert-info">
<strong>How would you create a list of dates in 2026 that are evenly spaced by 1 week?</strong>
<details><summary class='drop2'>Answer</summary>
```{r chald32bd}
date_list <- as.Date(c("01/01/2026", "12/31/2026"), format = "%m/%d/%Y") 
seq.Date(date_list[1], date_list[2], by = "1 week")
```
</details>
</div>

<div class="alert alert-info">
<strong>How would you extract the month as a number ranging from 1-12 in temp_data?.</strong>
<details><summary class='drop2'>Answer</summary>
```{r challdatec1}
temp_data$month_num <- as.numeric(format(temp_data$timestamp_temp, "%m"))
head(temp_data)
```
</details>
</div>

<div class="alert alert-info">
<strong>How would you extract the julian date in temp_data?.</strong>
<details><summary class='drop2'>Answer</summary>
```{r challdatejc1}
temp_data$julian <- as.numeric(format(temp_data$timestamp_temp, "%j"))
head(temp_data)
```
</details>
</div>

</details>
<br>