#### Day 3 Questions
<details open><summary class='drop'>Load Data and Packages</summary>
```{r}
library(dplyr)
library(stringr) # for word 
bat_cap <- read.csv("https://raw.githubusercontent.com/KateMMiller/IMD_R_Training_2026/refs/heads/main/data/example_bat_capture_data.csv")
```
</details><br>

<details open><summary class='drop'>Pivoting tables</summary>
<div class="alert alert-info">
<strong>Pivot the `bat_sum` data frame on year instead of species, so that you have a column for every year of captures. Remember to avoid column names starting with a number. </strong>

<details><summary class='code2'>Answer</summary>
```{r pivwide2c1, fold.output = F}
bat_wide_yr <- pivot_wider(bat_sum, names_from = Year, 
                           values_from = num_indiv, 
                           values_fill = 0, 
                           names_prefix = "yr")
head(bat_wide_yr)
```
</details>
</div>

<div class="alert alert-info">
<strong>Pivot the resulting data frame from the previous question to long on the years columns, and remove the "yr" from the year names using `names_prefix = 'yr'`. </strong>

<details><summary class='code2'>Answer</summary>
```{r pivwide2c2, fold.output = F}
bat_long_yr <- pivot_longer(bat_wide_yr, 
                            cols = -c(Site, sppcode),
                            names_to = "Year", 
                            values_to = "num_indiv", 
                            names_prefix = "yr") # drops this string from values
```
</details>
</div>
</details><br> 

<details open><summary class='drop'>Joining Tables</summary>





</details>
</br>