---
output: 
  html_document:
    css: custom_styles.css
---

#### Customizing ggplot
<details open><summary class = 'drop'>Load and prep data</summary>
For this section, we're going to use NETN water quality data to customize ggplot objects. This is an abbreviated dataset from the NETN water package. The data contains surface lake measurements recorded using a YSI in a subset of lakes in Acadia NP. We'll filter the data to plot at different combinations of parameters and sites. 

```{r chem, eval = F}
library(dplyr)
library(ggplot2)
chem <- read.csv("https://raw.githubusercontent.com/KateMMiller/IMD_R_Training_2026/refs/heads/main/data/NETN_water_chemistry_data.csv")
str(chem)
```
```{r chem2, echo = F}
chem <- read.csv("./data/NETN_water_chemistry_data.csv")
str(chem)
```

To start with, we're going to plot pH for 8 lakes monitored in Acadia NP. Before we start plotting, we need to convert the EventDate column to a date type, and extract a year, month, and day of year column for easier plotting. 

<p class ='code'>Filter chem data on sites and pH </p>
```{r chem3}
ACAD_lakes <- c("ACANTB", "ACBUBL", "ACEAGL", "ACECHO", 
                "ACJORD", "ACLONG", "ACSEAL", "ACWHOL")

lakes_pH <- chem |> filter(SiteCode %in% ACAD_lakes) |> 
  filter(Parameter == "pH") |> 
  mutate(date = as.Date(EventDate, "%m/%d/%Y"),
         year = as.numeric(format(date, "%Y")),
         mon = as.numeric(format(date, "%m")),
         doy = as.numeric(format(date, "%j")))
head(lakes_pH)
```
</details>
<br>

<details open><summary class = 'drop'>Custom palettes</summary>
Now that we have the data set up, we're going to make a line and point time series plot of pH for the different sites.

<p class = 'code'>Make a generic plot</p>
```{r polt1}
ggplot(lakes_pH, aes(x = doy, y = Value)) + theme_bw() +
  geom_point() 
```

#+++++ ENDED HERE ++++++

<p class = 'code'>Set color by SiteCode using default colors</p>
```{r polt2, eval = F}
ggplot(lakes_pH, aes(x = doy, y = Value, color = SiteCode)) + theme_bw() +
  geom_point() + 
  scale_color_manual()

```
The default colors in ggplot aren't great. Whenever you see these colors used in publications, you kind of know the author either barely knows ggplot or is lazy. Let's specify our own colors manually. 



</details>
<br>

<details open><summary class = 'drop'>Axes for dates</summary>


</details>
<br>
<details open><summary class = 'drop'>Facets and grids</summary>

</details>
<br>



<details open><summary class = 'drop'>Legends</summary>

</details>
<br>